Handling String-Number Conversion in String Constraint Solving

### 2019-11-05 draft.tex
some bug fix

\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\sti}[1]{\mbox{\textbf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textbf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvar}{\mbox{$\mathbb{V}_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{\mbox{$\mathbb{V}^1_{\Sigma_\epsilon}$}}
\newcommand{\cvartwo}{\mbox{$\mathbb{V}^2_{\Sigma_\epsilon}$}}
\newcommand{\modelof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}

@@

In this paper, we fix a finite alphabet $\Sigma \subseteq \mathbb{N}$. Note that here we assume the alphabet is a finite subset of natural numbers. Essentially, we try to capture the numerical encoding of the corresponding symbols in computers (e.g., in ASCII, `A' is encoded as $65$). We assume there is a one-one mapping between numbers in $\Sigma$ and the character it encodes. For simplicity of presentation, we assume the character `$0$' is mapped to number $0$, `$1$' to $1$,$\ldots$, and `$9$' to $9$. For other character $c$, we use $\enc{c}$ to denote the number that it maps to. Notice that this approach is general enough to support any finite set of characters. We write $[n,m]$ for the set of numbers (or symbols in $\Sigma$) $\{k\mid n\leq k \leq m\}$. A \emph{term} is an element in $(\vars\cup \Sigma)^*$.

Given a finite automaton $A$ over $\{a_1,a_2,\ldots,a_n\}$, the Parikh image of $L$ is the set $\{(|w|_{a_1},\ldots, |w|_{a_n}) \mid w \in  L(A) \}$, i.e., the number of occurrences of each symbol $\{a_1,a_2,\ldots,a_n\}$ of words in $L(A)$. Moreover, one can construct from $A$ in linear time a Presburger formula $\phi_A(v_1, \ldots,v_n)$ such that $\phi(c_1, \ldots,c_n) \iff (c_1, \ldots,c_n)$ in the Parikh image of $L$~\cite{peter}.

We use $\vars$ to denote a finite set of string variables ranging over $\Sigma^*$, $\varn$ to denote a finite set of integer variables ranging over $\mathbb{Z}$, and $\cvar$ to denote a finite set of character variables  ranging over $\Sigma_\epsilon$. Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. 
We call a string over $\cvar$ a \emph{symbolic string}. 
For a variable $x$, we call its primed version (e.g., $x'$) or indexed versions (e.g., $x_1$, $x^1$) its \emph{variants}.  
In this paper, we use $x,y$ or their variants to denote variables in $\vars$, $n$ or its variants to denote variables in $\varn$, and $v$ and its variants for variables in $\cvar$.

An \emph{interpretation over $\vars$ and $\varn$} is a mapping $I$ from $\vars\cup \varn$ to $\Sigma^* \cup \mathbb{N}$. We lift $i$ to terms by defining $I: (\vars\cup \Sigma)^* \rightarrow \Sigma^*$ with $I(a) =a$ for $a\in \Sigma$ and $I(t_1 \cdot t_2) = I(t_1)\cdot I(t_2)$.
An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{String Constraints} \label{section:sc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The \emph{string-number conversion function} $\sti{x}$ is defined as follows. For $a\in [0,9]$, we have $\sti{a}=a$ and for $x \cdot a \in [0,9]^+$, $\sti{x\cdot a} = 10\times \sti{x}+a$. For $x\notin [0,9]^+$, $\sti{x}=-1$.

Notice that only positive integer is supported in this semantics. This is the semantics used by most of the SMT solvers, and hence we follow it in this paper. The encoding has a benefit that only can also handle the case where $x$ is ``not a number'', using the condition $\sti{x} = -1$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Symbolic Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}

>>>>

\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\sti}[1]{\mbox{\textbf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textbf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{\mbox{$V^1_{\Sigma_\epsilon}$}}
\newcommand{\cvartwo}{\mbox{$V^2_{\Sigma_\epsilon}$}}
\newcommand{\modelof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}

@@

Given a finite automaton $A$ over $\{a_1,a_2,\ldots,a_n\}$, the Parikh image of $L$ is the set $\{(|w|_{a_1},\ldots, |w|_{a_n}) \mid w \in  L(A) \}$, i.e., the number of occurrences of each symbol $\{a_1,a_2,\ldots,a_n\}$ of words in $L(A)$. Moreover, one can construct from $A$ in linear time a Presburger formula $\phi_A(v_1, \ldots,v_n)$ such that $\phi(c_1, \ldots,c_n) \iff (c_1, \ldots,c_n)$ in the Parikh image of $L$~\cite{SeidlSMH04}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{String Constraints} \label{section:sc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section, we formally define string constraints and the decision problem we want to solve. We fix a finite alphabet $\Sigma \subseteq \mathbb{N}$. Note that here we assume the alphabet is a finite subset of natural numbers. Essentially, we try to capture the numerical encoding of the corresponding symbols in computers (e.g., in ASCII, `A' is encoded as $65$). We assume there is a one-one mapping between numbers in $\Sigma$ and the character it encodes. For simplicity of presentation, we assume the character `$0$' is mapped to number $0$, `$1$' to $1$,$\ldots$, and `$9$' to $9$. For other character $c$, we use $\enc{c}$ to denote the number that it maps to. Notice that this approach is general enough to support any finite set of characters. We write $[n,m]$ for the set of numbers (or symbols in $\Sigma$) $\{k\mid n\leq k \leq m\}$. A \emph{term} is an element in $(\vars\cup \Sigma)^*$.

We call $\vars$ the set of \emph{string variables} (ranging over $\Sigma^*$) and $\varn$ the set of \emph{integer variables} (ranging over $\mathbb{Z}$).
For a variable $x$, we call its primed version (e.g., $x'$) or indexed versions (e.g., $x_1$, $x^1$) its \emph{variants}.
In this paper, we use $x,y$ (or their variants) to denote string variables and $n$ (or its variants) to denote integer variables.

The \emph{string-number conversion function} $\sti{x}$ is defined as follows. For $a\in [0,9]$, we have $\sti{a}=a$ and for $x \cdot a \in [0,9]^+$, $\sti{x\cdot a} = 10\times \sti{x}+a$. For $x\notin [0,9]^+$, $\sti{x}=-1$.

Notice that only positive integer is supported in this semantics. This is the semantics used by most of the SMT solvers, and hence we follow it in this paper. The encoding has a benefit that it can also handle the case where $x$ is ``not a number'', using the condition $\sti{x} = -1$.

@@

An \emph{interpretation over $\vars$ and $\varn$} is a mapping $I$ from $\vars\cup \varn$ to $\Sigma^* \cup \mathbb{N}$. We lift $I$ to terms (i.e., an element in $(\vars\cup \Sigma)^*$) in the standard manner.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Symbolic Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}
We use $\cvar$ to denote a finite set of \emph{character variables}  ranging over $\Sigma_\epsilon$. Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. We use $v$ and its variants for character variables. We call a string over $\cvar$ (character variables) a \emph{symbolic string}.
An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.
